---
title: "Data"
bibliography: references.bib
csl: apa.csl
---

```{r, message=FALSE, warning=FALSE, include=FALSE}

# load packages used

library(readxl)
library(tidyverse)
library(glmnet)
library(randomForest)
library(caret)
library(car)
library(knitr)
library(patchwork)
library(leaps)

```

Over the years 2020-2024, the CDC’s State Unintentional Drug Overdose Reporting System (SUDORS) compiled data on unintentional and undetermined-intent drug overdose deaths, stratified by drug type, age, and sex, in 33-43 states including DC. A portion of this Fatal Drug Overdose Data [@cdc_sudors] was combined with income and educational attainment data from the Federal Reserve Bank of St. Louis (FRED) [@fred_income; @fred_hs; @fred_bachelors] and poverty data from the US Census Bureau [@census_poverty2022; @census_poverty2023; @census_poverty2024]. There are 187 observations.

```{r, message=FALSE, warning=FALSE, results='hide', echo=FALSE, include=FALSE}

# load data
od_data = read_excel("overdose_data.xlsx")

glimpse(od_data)

# change state and year to factor variables
od_data = od_data %>%
  mutate(
    state = factor(state),
    year = factor(year),
    med_income = med_income/1000
  )

# change all 9999 values to NA
od_data = od_data %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))

```

We explore how the following variables can be used to explain and predict overdose death rates, given as an age-adjusted rate per 100,000 persons, for any/all drugs, heroin, prescription opioids, cocaine, methamphetamine, and benzodiazepines.

```{r,  message=FALSE, warning=FALSE, echo=FALSE}

vars = data.frame(
  Variable = c("`median_income` ", "`poverty`", "`hs_grad`", "`bachelors`"),
  Description = c(
    "median household income in thousands (USD)",
    "percentage of the population below the national poverty threshold (i.e., poverty rate)",
    "percentage of the population 18 years and over with a high school diploma or higher",
    "percentage of the population 18 years and over with a bachelor's degree or higher"
  )
)

kable(vars, col.names = NULL)
```

The following graphs show the distribution of overdose death rates for each drug. We note that all of the distributions are unimodal and right-skewed.

```{r,  message=FALSE, warning=FALSE, echo=FALSE}

od_long = od_data %>%
  select(alldrug_rate, heroin_rate, rxopioids_rate, 
         cocaine_rate, meth_rate, benzodiazepines_rate) %>%
  pivot_longer(
    cols = everything(),
    names_to = "drug",
    values_to = "rate"
  )

od_long$drug = factor(od_long$drug,
  levels = c("alldrug_rate", "heroin_rate", "rxopioids_rate",
             "cocaine_rate", "meth_rate", "benzodiazepines_rate"),
  labels = c("All Drugs", "Heroin", "Prescription Opioids",
             "Cocaine", "Methamphetamine", "Benzodiazepines")
)

ggplot(od_long, aes(x = rate)) +
  geom_histogram(bins = 20, color = "black", fill = "lightsteelblue2") +
  facet_wrap(~ drug, scales = "free_x") +
  labs(
    x = "Overdose Death Rate (per 100,000 persons)",
    y = "Count",
    title = "Distribution of Overdose Death Rates by Drug Type"
  )

```

We also present scatterplots illustrating relationships between each predictor variable and the overdose death rate for each drug type.

## All drugs:

No clear relationship can be discerned between states' drug overdose death rates and median household income, proportion of the population with a high school diploma or higher, or proportion of the population with a bachelor's degree or higher. There appears to be a weak positive correlation between drug overdose deaths and poverty rate ($r = 0.33$).

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = alldrug_rate)) +
      geom_point(color = "steelblue") +
      labs(
       x = "Median Income (in thousands)",
       y = "Overdose Death Rate",
       title = "Overdose Deaths vs Median Income"
     )


p2 = ggplot(od_data, aes(x = poverty, y = alldrug_rate)) +
      geom_point(color = "steelblue") +
      labs(
        x = "Poverty Rate",
        y = "Overdose Death Rate",
        title = "Overdose Deaths vs Poverty Rate"
     )

p1 + p2


edu_long = od_data %>%
  select(alldrug_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = alldrug_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# correlation between alldrug_rate and poverty
d = od_data %>%
  select(alldrug_rate, poverty) %>%
  na.omit()
cor(d$alldrug_rate, d$poverty)

```

## Heroin:

No clear relationship can be discerned between states' heroin overdose death rates and median household income, poverty rate, proportion of the population with a high school diploma or higher, or proportion of the population with a bachelor's degree or higher.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = heroin_rate)) +
      geom_point(color = "steelblue") +
      labs(
       x = "Median Income (in thousands)",
       y = "Overdose Death Rate",
       title = "Overdose Deaths vs Median Income"
     )


p2 = ggplot(od_data, aes(x = poverty, y = heroin_rate)) +
      geom_point(color = "steelblue") +
      labs(
        x = "Poverty Rate",
        y = "Overdose Death Rate",
        title = "Overdose Deaths vs Poverty Rate"
     )

p1 + p2


edu_long = od_data %>%
  select(heroin_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = heroin_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

## Prescription Opioids:

Omitting a few outliers, no clear relationship can be discerned between states' prescription opioid overdose death rates and median household income, poverty rate, proportion of the population with a high school diploma or higher, or proportion of the population with a bachelor's degree or higher.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(rxopioids_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

## Cocaine:

No clear relationship can be discerned between states' cocaine overdose death rates and median household income, poverty rate, proportion of the population with a high school diploma or higher. There appears to be a moderate positive correlation between cocaine overdose deaths and proportion of the population with a bachelors degree or higher ($r = 0.67$).

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(cocaine_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# compute correlation coefficient between cocaine_rate and bachelors
c = od_data %>%
    select(cocaine_rate, bachelors) %>%
    drop_na()
cor(c$cocaine_rate, c$bachelors)

```

## Methamphetamine:

There appears to be a moderate-to-weak negative correlation between states' methamphetamine overdose death rates and median household income ($r = -0.45$) and a moderate-to-weak positive correlation between death rates and poverty rate ($r = 0.48$). Omitting a few outliers, no clear relationship can be discerned between methamphetamine overdose deaths and proportion of the population with a high school diploma or higher, but a moderate-to-strong negative relationship appears to exist between the death rate and proportion of the population holding a bachelors degree or higher.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = meth_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = meth_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(meth_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = meth_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# compute correlation coefficient between meth_rate and med_income
m = od_data %>%
    select(meth_rate, med_income) %>%
    drop_na()
cor(m$meth_rate, m$med_income)

# compute correlation coefficient between meth_rate and poverty
m = od_data %>%
    select(meth_rate, poverty) %>%
    drop_na()
cor(m$meth_rate, m$poverty)

```

## Benzodiazepines:

Omitting a few outliers, no clear relationship can be discerned between states' drug overdose death rates and median household income, poverty rate, proportion of the population with a high school diploma or higher, or proportion of the population with a bachelor's degree or higher.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(benzodiazepines_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```
