---
title: "R Code"
---

```{r, eval=FALSE}

# load packages used

library(readxl)
library(tidyverse)
library(glmnet)
library(randomForest)
library(caret)
library(car)
library(knitr)
library(patchwork)
library(leaps)

```

# Data

```{r, eval=FALSE}

# load data
od_data = read_excel("overdose_data.xlsx")

glimpse(od_data)

# change state and year to factor variables
od_data = od_data %>%
  mutate(
    state = factor(state),
    year = factor(year),
    med_income = med_income/1000
  )

# change all 9999 values to NA
od_data = od_data %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))

```

```{r, eval=FALSE}

vars = data.frame(
  Variable = c("`median_income` ", "`poverty`", "`hs_grad`", "`bachelors`"),
  Description = c(
    "median household income in thousands (USD)",
    "percentage of the population below the national poverty threshold (i.e., poverty rate)",
    "percentage of the population 18 years and over with a high school diploma or higher",
    "percentage of the population 18 years and over with a bachelor's degree or higher"
  )
)

kable(vars, col.names = NULL)
```

```{r, eval=FALSE}

od_long = od_data %>%
  select(alldrug_rate, heroin_rate, rxopioids_rate, 
         cocaine_rate, meth_rate, benzodiazepines_rate) %>%
  pivot_longer(
    cols = everything(),
    names_to = "drug",
    values_to = "rate"
  )

od_long$drug = factor(od_long$drug,
  levels = c("alldrug_rate", "heroin_rate", "rxopioids_rate",
             "cocaine_rate", "meth_rate", "benzodiazepines_rate"),
  labels = c("All Drugs", "Heroin", "Prescription Opioids",
             "Cocaine", "Methamphetamine", "Benzodiazepines")
)

ggplot(od_long, aes(x = rate)) +
  geom_histogram(bins = 20, color = "black", fill = "lightsteelblue2") +
  facet_wrap(~ drug, scales = "free_x") +
  labs(
    x = "Overdose Death Rate (per 100,000 persons)",
    y = "Count",
    title = "Distribution of Overdose Death Rates by Drug Type"
  )

```

## All drugs:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = alldrug_rate)) +
      geom_point(color = "steelblue") +
      labs(
       x = "Median Income (in thousands)",
       y = "Overdose Death Rate",
       title = "Overdose Deaths vs Median Income"
     )


p2 = ggplot(od_data, aes(x = poverty, y = alldrug_rate)) +
      geom_point(color = "steelblue") +
      labs(
        x = "Poverty Rate",
        y = "Overdose Death Rate",
        title = "Overdose Deaths vs Poverty Rate"
     )

p1 + p2


edu_long = od_data %>%
  select(alldrug_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = alldrug_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, eval=FALSE}

# correlation between alldrug_rate and poverty
d = od_data %>%
  select(alldrug_rate, poverty) %>%
  na.omit()
cor(d$alldrug_rate, d$poverty)

```

## Heroin:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = heroin_rate)) +
      geom_point(color = "steelblue") +
      labs(
       x = "Median Income (in thousands)",
       y = "Overdose Death Rate",
       title = "Overdose Deaths vs Median Income"
     )


p2 = ggplot(od_data, aes(x = poverty, y = heroin_rate)) +
      geom_point(color = "steelblue") +
      labs(
        x = "Poverty Rate",
        y = "Overdose Death Rate",
        title = "Overdose Deaths vs Poverty Rate"
     )

p1 + p2


edu_long = od_data %>%
  select(heroin_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = heroin_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

## Prescription Opioids:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(rxopioids_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = rxopioids_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

## Cocaine:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(cocaine_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = cocaine_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, eval=FALSE}

# compute correlation coefficient between cocaine_rate and bachelors
c = od_data %>%
    select(cocaine_rate, bachelors) %>%
    drop_na()
cor(c$cocaine_rate, c$bachelors)

```

## Methamphetamine:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = meth_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = meth_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(meth_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = meth_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

```{r, eval=FALSE}

# compute correlation coefficient between meth_rate and med_income
m = od_data %>%
    select(meth_rate, med_income) %>%
    drop_na()
cor(m$meth_rate, m$med_income)

# compute correlation coefficient between meth_rate and poverty
m = od_data %>%
    select(meth_rate, poverty) %>%
    drop_na()
cor(m$meth_rate, m$poverty)

```

## Benzodiazepines:

```{r, eval=FALSE}

p1 = ggplot(od_data, aes(x = med_income, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Median Income (in thousands)",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Median Income"
  )


p2 = ggplot(od_data, aes(x = poverty, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  labs(
    x = "Poverty Rate",
    y = "Overdose Death Rate",
    title = "Overdose Deaths vs Poverty Rate"
  )

p1 + p2


edu_long = od_data %>%
  select(benzodiazepines_rate, hs_grad, bachelors) %>%
  pivot_longer(
    cols = c(hs_grad, bachelors),
    names_to = "education_level",
    values_to = "percent"
  )

edu_long$education_level = factor(
  edu_long$education_level,
  levels = c("hs_grad", "bachelors"),
  labels = c("High School Diploma or Higher", "Bachelor's Degree or Higher")
)

ggplot(edu_long, aes(x = percent, y = benzodiazepines_rate)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ education_level, scales = "free_x") +
  labs(
    x = "Percent of Population",
    y = "Overdose Death Rate",
    title = "Overdose Death Rate vs. Educational Attainment"
  )

```

# Research Question 1: How are median household income, poverty rate, and educational attainment related to drug overdose death rates in states, and do these relationships vary across different drugs?

## Results

### All drugs

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(alldrug_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$alldrug_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
alldrug_model_hs = lm(alldrug_rate ~ med_income + poverty + hs_grad, 
                       data = data_hs)

summary(alldrug_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(alldrug_rate ~ med_income + poverty + hs_grad,
             data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$alldrug_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

alldrug_hs.mse = min(mean.cv.errors)
alldrug_hs.rmse = sqrt(alldrug_hs.mse)
alldrug_hs.mae = mean(cv.mae)
alldrug_hs.adjr2 = summary(alldrug_model_hs)$adj.r.squared

cat("Test MSE =", alldrug_hs.mse, "\n")
cat("Test RMSE =", alldrug_hs.rmse, "\n")
cat("Test MAE =", alldrug_hs.mae, "\n")
cat("Adjusted R-squared =", alldrug_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(alldrug_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
#                       data = data_bach, 
#                       method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$alldrug_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
                         data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
alldrug_model_bach = lm(alldrug_rate ~ poverty + bachelors, 
                       data = data_bach)


summary(alldrug_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(alldrug_rate ~ poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$alldrug_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

alldrug_bach.mse = min(mean.cv.errors)
alldrug_bach.rmse = sqrt(alldrug_bach.mse)
alldrug_bach.mae = mean(cv.mae)
alldrug_bach.adjr2 = summary(alldrug_model_bach)$adj.r.squared

cat("Test MSE =", alldrug_bach.mse, "\n")
cat("Test RMSE =", alldrug_bach.rmse, "\n")
cat("Test MAE =", alldrug_bach.mae, "\n")
cat("Adjusted R-squared =", alldrug_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(alldrug_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
alldrug_hs.ridge.mse = mean((actual - ridge.pred)^2)
alldrug_hs.ridge.rmse = sqrt(alldrug_hs.ridge.mse)
alldrug_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
alldrug_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
                sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

alldrug_hs.ridge.adjr2 = 1 - ((1 - alldrug_hs.ridge.r2) * (n - 1)) /
                    (n - p - 1)


cat("MSE =", alldrug_hs.ridge.mse, "\n")
cat("RMSE =", alldrug_hs.ridge.rmse, "\n")
cat("MAE =", alldrug_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", alldrug_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(alldrug_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.alldrug = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.alldrug = y

# MSE, RMSE, MAE
alldrug_bach.ridge.mse = mean((actual.alldrug - ridge.pred.alldrug)^2)
alldrug_bach.ridge.rmse = sqrt(alldrug_bach.ridge.mse)
alldrug_bach.ridge.mae = mean(abs(actual.alldrug - ridge.pred.alldrug))

# R-squared
alldrug_bach.ridge.r2 = 1 - sum((actual.alldrug - ridge.pred.alldrug)^2) /
  sum((actual.alldrug - mean(actual.alldrug))^2)

# Adjusted R-squared
n = length(actual.alldrug)
p = ncol(x)

alldrug_bach.ridge.adjr2 = 1 - ((1 - alldrug_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", alldrug_bach.ridge.mse, "\n")
cat("RMSE =", alldrug_bach.ridge.rmse, "\n")
cat("MAE =", alldrug_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", alldrug_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
alldrug_baseline.pred = mean(od_data$alldrug_rate, na.rm = TRUE)

alldrug_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


alldrug_baseline.rmse = sqrt(mean((od_data$alldrug_rate - alldrug_baseline.pred)^2,
                          na.rm = TRUE))

alldrug_baseline.mae = mean(abs(od_data$alldrug_rate - alldrug_baseline.pred.med),
                    na.rm = TRUE)

alldrug_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(alldrug_baseline.rmse,
           alldrug_hs.rmse,
           alldrug_bach.rmse,
           alldrug_hs.ridge.rmse,
           alldrug_bach.ridge.rmse),
  
  MAE = c(alldrug_baseline.mae,
          alldrug_hs.mae,
          alldrug_bach.mae,
          alldrug_hs.ridge.mae,
          alldrug_bach.ridge.mae),
  
  Adjusted_R2 = c(alldrug_baseline.adjr2,
                  alldrug_hs.adjr2,
                  alldrug_bach.adjr2,
                  alldrug_hs.ridge.adjr2,
                  alldrug_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = '"All Drugs" Model Comparison')

```

### Heroin

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(heroin_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$heroin_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
heroin_model_hs = lm(heroin_rate ~ med_income + poverty + hs_grad, 
                      data = data_hs)

summary(heroin_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(heroin_rate ~ med_income + poverty + hs_grad,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$heroin_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

heroin_hs.mse = min(mean.cv.errors)
heroin_hs.rmse = sqrt(heroin_hs.mse)
heroin_hs.mae = mean(cv.mae)
heroin_hs.adjr2 = summary(heroin_model_hs)$adj.r.squared

cat("Test MSE =", heroin_hs.mse, "\n")
cat("Test RMSE =", heroin_hs.rmse, "\n")
cat("Test MAE =", heroin_hs.mae, "\n")
cat("Adjusted R-squared =", heroin_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(heroin_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$heroin_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
heroin_model_bach = lm(heroin_rate ~ med_income + poverty + bachelors, 
                        data = data_bach)


summary(heroin_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(heroin_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$heroin_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

heroin_bach.mse = min(mean.cv.errors)
heroin_bach.rmse = sqrt(heroin_bach.mse)
heroin_bach.mae = mean(cv.mae)
heroin_bach.adjr2 = summary(heroin_model_bach)$adj.r.squared

cat("Test MSE =", heroin_bach.mse, "\n")
cat("Test RMSE =", heroin_bach.rmse, "\n")
cat("Test MAE =", heroin_bach.mae, "\n")
cat("Adjusted R-squared =", heroin_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(heroin_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.heroin = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.heroin = y

# MSE, RMSE, MAE
heroin_hs.ridge.mse = mean((actual.heroin - ridge.pred.heroin)^2)
heroin_hs.ridge.rmse = sqrt(heroin_hs.ridge.mse)
heroin_hs.ridge.mae = mean(abs(actual.heroin - ridge.pred.heroin))

# R-squared
heroin_hs.ridge.r2 = 1 - sum((actual.heroin - ridge.pred.heroin)^2) /
  sum((actual.heroin - mean(actual.heroin))^2)

# Adjusted R-squared
n = length(actual.heroin)
p = ncol(x)

heroin_hs.ridge.adjr2 = 1 - ((1 - heroin_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_hs.ridge.mse, "\n")
cat("RMSE =", heroin_hs.ridge.rmse, "\n")
cat("MAE =", heroin_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", heroin_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(heroin_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
heroin_bach.ridge.mse = mean((actual - ridge.pred)^2)
heroin_bach.ridge.rmse = sqrt(heroin_bach.ridge.mse)
heroin_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
heroin_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

heroin_bach.ridge.adjr2 = 1 - ((1 - heroin_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_bach.ridge.mse, "\n")
cat("RMSE =", heroin_bach.ridge.rmse, "\n")
cat("MAE =", heroin_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", heroin_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
heroin_baseline.pred = mean(od_data$heroin_rate, na.rm = TRUE)

heroin_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


heroin_baseline.rmse = sqrt(mean((od_data$heroin_rate - heroin_baseline.pred)^2,
                                  na.rm = TRUE))

heroin_baseline.mae = mean(abs(od_data$heroin_rate - heroin_baseline.pred.med),
                            na.rm = TRUE)

heroin_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(heroin_baseline.rmse,
           heroin_hs.rmse,
           heroin_bach.rmse,
           heroin_hs.ridge.rmse,
           heroin_bach.ridge.rmse),
  
  MAE = c(heroin_baseline.mae,
          heroin_hs.mae,
          heroin_bach.mae,
          heroin_hs.ridge.mae,
          heroin_bach.ridge.mae),
  
  Adjusted_R2 = c(heroin_baseline.adjr2,
                  heroin_hs.adjr2,
                  heroin_bach.adjr2,
                  heroin_hs.ridge.adjr2,
                  heroin_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Heroin Model Comparison')

```

### Prescription Opioids

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(rxopioids_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$rxopioids_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")

coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
rxopioids_model_hs = lm(rxopioids_rate ~ med_income + poverty, 
                     data = data_hs)

summary(rxopioids_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(rxopioids_rate ~ med_income + poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$rxopioids_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

rxopioids_hs.mse = min(mean.cv.errors)
rxopioids_hs.rmse = sqrt(rxopioids_hs.mse)
rxopioids_hs.mae = mean(cv.mae)
rxopioids_hs.adjr2 = summary(rxopioids_model_hs)$adj.r.squared

cat("Test MSE =", rxopioids_hs.mse, "\n")
cat("Test RMSE =", rxopioids_hs.rmse, "\n")
cat("Test MAE =", rxopioids_hs.mae, "\n")
cat("Adjusted R-squared =", rxopioids_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(rxopioids_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$rxopioids_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
rxopioids_model_bach = lm(rxopioids_rate ~ poverty + bachelors, 
                       data = data_bach)


summary(rxopioids_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(rxopioids_rate ~ poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$rxopioids_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

rxopioids_bach.mse = min(mean.cv.errors)
rxopioids_bach.rmse = sqrt(rxopioids_bach.mse)
rxopioids_bach.mae = mean(cv.mae)
rxopioids_bach.adjr2 = summary(rxopioids_model_bach)$adj.r.squared

cat("Test MSE =", rxopioids_bach.mse, "\n")
cat("Test RMSE =", rxopioids_bach.rmse, "\n")
cat("Test MAE =", rxopioids_bach.mae, "\n")
cat("Adjusted R-squared =", rxopioids_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(rxopioids_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_hs.ridge.mse = mean((actual - ridge.pred)^2)
rxopioids_hs.ridge.rmse = sqrt(rxopioids_hs.ridge.mse)
rxopioids_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_hs.ridge.adjr2 = 1 - ((1 - rxopioids_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_hs.ridge.mse, "\n")
cat("RMSE =", rxopioids_hs.ridge.rmse, "\n")
cat("MAE =", rxopioids_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", rxopioids_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(rxopioids_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_bach.ridge.mse = mean((actual - ridge.pred)^2)
rxopioids_bach.ridge.rmse = sqrt(rxopioids_bach.ridge.mse)
rxopioids_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_bach.ridge.adjr2 = 1 - ((1 - rxopioids_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_bach.ridge.mse, "\n")
cat("RMSE =", rxopioids_bach.ridge.rmse, "\n")
cat("MAE =", rxopioids_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", rxopioids_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
rxopioids_baseline.pred = mean(od_data$rxopioids_rate, na.rm = TRUE)

rxopioids_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


rxopioids_baseline.rmse = sqrt(mean((od_data$rxopioids_rate - rxopioids_baseline.pred)^2,
                                 na.rm = TRUE))

rxopioids_baseline.mae = mean(abs(od_data$rxopioids_rate - rxopioids_baseline.pred.med),
                           na.rm = TRUE)

rxopioids_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(rxopioids_baseline.rmse,
           rxopioids_hs.rmse,
           rxopioids_bach.rmse,
           rxopioids_hs.ridge.rmse,
           rxopioids_bach.ridge.rmse),
  
  MAE = c(rxopioids_baseline.mae,
          rxopioids_hs.mae,
          rxopioids_bach.mae,
          rxopioids_hs.ridge.mae,
          rxopioids_bach.ridge.mae),
  
  Adjusted_R2 = c(rxopioids_baseline.adjr2,
                  rxopioids_hs.adjr2,
                  rxopioids_bach.adjr2,
                  rxopioids_hs.ridge.adjr2,
                  rxopioids_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Prescription Opioids Model Comparison')

```

### Cocaine

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(cocaine_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$cocaine_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
cocaine_model_hs = lm(cocaine_rate ~ med_income + poverty + hs_grad, 
                     data = data_hs)

summary(cocaine_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(cocaine_rate ~ med_income + poverty + hs_grad,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$cocaine_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

cocaine_hs.mse = min(mean.cv.errors)
cocaine_hs.rmse = sqrt(cocaine_hs.mse)
cocaine_hs.mae = mean(cv.mae)
cocaine_hs.adjr2 = summary(cocaine_model_hs)$adj.r.squared

cat("Test MSE =", cocaine_hs.mse, "\n")
cat("Test RMSE =", cocaine_hs.rmse, "\n")
cat("Test MAE =", cocaine_hs.mae, "\n")
cat("Adjusted R-squared =", cocaine_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(cocaine_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$cocaine_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
cocaine_model_bach = lm(cocaine_rate ~ med_income + bachelors, 
                       data = data_bach)


summary(cocaine_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(cocaine_rate ~ med_income + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$cocaine_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

cocaine_bach.mse = min(mean.cv.errors)
cocaine_bach.rmse = sqrt(cocaine_bach.mse)
cocaine_bach.mae = mean(cv.mae)
cocaine_bach.adjr2 = summary(cocaine_model_bach)$adj.r.squared

cat("Test MSE =", cocaine_bach.mse, "\n")
cat("Test RMSE =", cocaine_bach.rmse, "\n")
cat("Test MAE =", cocaine_bach.mae, "\n")
cat("Adjusted R-squared =", cocaine_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(cocaine_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_hs.ridge.mse = mean((actual - ridge.pred)^2)
cocaine_hs.ridge.rmse = sqrt(cocaine_hs.ridge.mse)
cocaine_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_hs.ridge.adjr2 = 1 - ((1 - cocaine_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", cocaine_hs.ridge.mse, "\n")
cat("RMSE =", cocaine_hs.ridge.rmse, "\n")
cat("MAE =", cocaine_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", cocaine_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(cocaine_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_bach.ridge.mse = mean((actual - ridge.pred)^2)
cocaine_bach.ridge.rmse = sqrt(cocaine_bach.ridge.mse)
cocaine_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_bach.ridge.adjr2 = 1 - ((1 - cocaine_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", cocaine_bach.ridge.mse, "\n")
cat("RMSE =", cocaine_bach.ridge.rmse, "\n")
cat("MAE =", cocaine_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", cocaine_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
cocaine_baseline.pred = mean(od_data$cocaine_rate, na.rm = TRUE)

cocaine_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)

cocaine_baseline.rmse = sqrt(mean((od_data$cocaine_rate - cocaine_baseline.pred)^2,
                                 na.rm = TRUE))

cocaine_baseline.mae = mean(abs(od_data$cocaine_rate - cocaine_baseline.pred.med),
                           na.rm = TRUE)

cocaine_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(cocaine_baseline.rmse,
           cocaine_hs.rmse,
           cocaine_bach.rmse,
           cocaine_hs.ridge.rmse,
           cocaine_bach.ridge.rmse),
  
  MAE = c(cocaine_baseline.mae,
          cocaine_hs.mae,
          cocaine_bach.mae,
          cocaine_hs.ridge.mae,
          cocaine_bach.ridge.mae),
  
  Adjusted_R2 = c(cocaine_baseline.adjr2,
                  cocaine_hs.adjr2,
                  cocaine_bach.adjr2,
                  cocaine_hs.ridge.adjr2,
                  cocaine_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Cocaine Model Comparison')

```

### Methamphetamine

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(meth_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$meth_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
meth_model_hs = lm(meth_rate ~ poverty, 
                     data = data_hs)

summary(meth_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(meth_rate ~ poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$meth_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

meth_hs.mse = min(mean.cv.errors)
meth_hs.rmse = sqrt(meth_hs.mse)
meth_hs.mae = mean(cv.mae)
meth_hs.adjr2 = summary(meth_model_hs)$adj.r.squared

cat("Test MSE =", meth_hs.mse, "\n")
cat("Test RMSE =", meth_hs.rmse, "\n")
cat("Test MAE =", meth_hs.mae, "\n")
cat("Adjusted R-squared =", meth_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(meth_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$meth_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
meth_model_bach = lm(meth_rate ~ med_income + poverty + bachelors, 
                       data = data_bach)


summary(meth_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(meth_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$meth_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

meth_bach.mse = min(mean.cv.errors)
meth_bach.rmse = sqrt(meth_bach.mse)
meth_bach.mae = mean(cv.mae)
meth_bach.adjr2 = summary(meth_model_bach)$adj.r.squared

cat("Test MSE =", meth_bach.mse, "\n")
cat("Test RMSE =", meth_bach.rmse, "\n")
cat("Test MAE =", meth_bach.mae, "\n")
cat("Adjusted R-squared =", meth_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(meth_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
meth_hs.ridge.mse = mean((actual - ridge.pred)^2)
meth_hs.ridge.rmse = sqrt(meth_hs.ridge.mse)
meth_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
meth_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

meth_hs.ridge.adjr2 = 1 - ((1 - meth_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_hs.ridge.mse, "\n")
cat("RMSE =", meth_hs.ridge.rmse, "\n")
cat("MAE =", meth_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", meth_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(meth_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.meth = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.meth = y

# MSE, RMSE, MAE
meth_bach.ridge.mse = mean((actual.meth - ridge.pred.meth)^2)
meth_bach.ridge.rmse = sqrt(meth_bach.ridge.mse)
meth_bach.ridge.mae = mean(abs(actual.meth - ridge.pred.meth))

# R-squared
meth_bach.ridge.r2 = 1 - sum((actual.meth - ridge.pred.meth)^2) /
  sum((actual.meth - mean(actual.meth))^2)

# Adjusted R-squared
n = length(actual.meth)
p = ncol(x)

meth_bach.ridge.adjr2 = 1 - ((1 - meth_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_bach.ridge.mse, "\n")
cat("RMSE =", meth_bach.ridge.rmse, "\n")
cat("MAE =", meth_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", meth_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
meth_baseline.pred = mean(od_data$meth_rate, na.rm = TRUE)

meth_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


meth_baseline.rmse = sqrt(mean((od_data$meth_rate - meth_baseline.pred)^2,
                                 na.rm = TRUE))

meth_baseline.mae = mean(abs(od_data$meth_rate - meth_baseline.pred.med),
                           na.rm = TRUE)

meth_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(meth_baseline.rmse,
           meth_hs.rmse,
           meth_bach.rmse,
           meth_hs.ridge.rmse,
           meth_bach.ridge.rmse),
  
  MAE = c(meth_baseline.mae,
          meth_hs.mae,
          meth_bach.mae,
          meth_hs.ridge.mae,
          meth_bach.ridge.mae),
  
  Adjusted_R2 = c(meth_baseline.adjr2,
                  meth_hs.adjr2,
                  meth_bach.adjr2,
                  meth_hs.ridge.adjr2,
                  meth_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Methamphetamine Model Comparison')

```

### Benzodiazepines

```{r, eval=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$benzodiazepines_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
benzo_model_hs = lm(benzodiazepines_rate ~ poverty, 
                     data = data_hs)

summary(benzo_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(benzodiazepines_rate ~ poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$benzodiazepines_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

benzo_hs.mse = min(mean.cv.errors)
benzo_hs.rmse = sqrt(benzo_hs.mse)
benzo_hs.mae = mean(cv.mae)
benzo_hs.adjr2 = summary(benzo_model_hs)$adj.r.squared

cat("Test MSE =", benzo_hs.mse, "\n")
cat("Test RMSE =", benzo_hs.rmse, "\n")
cat("Test MAE =", benzo_hs.mae, "\n")
cat("Adjusted R-squared =", benzo_hs.adjr2, "\n")

```

```{r, eval=FALSE}

# data cleaning
data_bach = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$benzodiazepines_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
benzo_model_bach = lm(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                       data = data_bach)


summary(benzo_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(benzodiazepines_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$benzodiazepines_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

benzo_bach.mse = min(mean.cv.errors)
benzo_bach.rmse = sqrt(benzo_bach.mse)
benzo_bach.mae = mean(cv.mae)
benzo_bach.adjr2 = summary(benzo_model_bach)$adj.r.squared

cat("Test MSE =", benzo_bach.mse, "\n")
cat("Test RMSE =", benzo_bach.rmse, "\n")
cat("Test MAE =", benzo_bach.mae, "\n")
cat("Adjusted R-squared =", benzo_bach.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge hs

x = model.matrix(benzodiazepines_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_hs.ridge.mse = mean((actual - ridge.pred)^2)
benzo_hs.ridge.rmse = sqrt(benzo_hs.ridge.mse)
benzo_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
benzo_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

benzo_hs.ridge.adjr2 = 1 - ((1 - benzo_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_hs.ridge.mse, "\n")
cat("RMSE =", benzo_hs.ridge.rmse, "\n")
cat("MAE =", benzo_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", benzo_hs.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

# ridge bachelors

x = model.matrix(benzodiazepines_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_bach.ridge.mse = mean((actual - ridge.pred)^2)
benzo_bach.ridge.rmse = sqrt(benzo_bach.ridge.mse)
benzo_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
benzo_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

benzo_bach.ridge.adjr2 = 1 - ((1 - benzo_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_bach.ridge.mse, "\n")
cat("RMSE =", benzo_bach.ridge.rmse, "\n")
cat("MAE =", benzo_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", benzo_bach.ridge.adjr2, "\n")

```

```{r, eval=FALSE}

### stats comparisons

# baseline model
benzo_baseline.pred = mean(od_data$benzodiazepines_rate, na.rm = TRUE)
benzo_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


benzo_baseline.rmse = sqrt(mean((od_data$benzodiazepines_rate - benzo_baseline.pred)^2,
                                 na.rm = TRUE))

benzo_baseline.mae = mean(abs(od_data$benzodiazepines_rate - benzo_baseline.pred.med),
                           na.rm = TRUE)

benzo_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(benzo_baseline.rmse,
           benzo_hs.rmse,
           benzo_bach.rmse,
           benzo_hs.ridge.rmse,
           benzo_bach.ridge.rmse),
  
  MAE = c(benzo_baseline.mae,
          benzo_hs.mae,
          benzo_bach.mae,
          benzo_hs.ridge.mae,
          benzo_bach.ridge.mae),
  
  Adjusted_R2 = c(benzo_baseline.adjr2,
                  benzo_hs.adjr2,
                  benzo_bach.adjr2,
                  benzo_hs.ridge.adjr2,
                  benzo_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Benzodiazepines Model Comparison')

```

# Research Question 2: Does the effect of poverty on overdose death rates depend on educational attainment of the population?

## Results

### All drugs

```{r, eval=FALSE}

# data cleaning
data_alldrug = od_data %>%
  select(alldrug_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(alldrug_rate ~ med_income + poverty*bachelors,
                 data = data_alldrug)[, -1]

y = data_alldrug$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
alldrug_int.mse = mean((actual - ridge.pred)^2)
alldrug_int.rmse = sqrt(alldrug_int.mse)
alldrug_int.mae = mean(abs(actual - ridge.pred))

# R-squared
alldrug_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

alldrug_int.adjr2 = 1 - ((1 - alldrug_int.r2) * (n - 1)) /
  (n - p - 1)

cat("MSE =", alldrug_int.mse, "\n")
cat("RMSE =", alldrug_int.rmse, "\n")
cat("MAE =", alldrug_int.mae, "\n")
cat("Adjusted R-squared =", alldrug_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(alldrug_int.rmse,
           alldrug_bach.ridge.rmse),
  
  MAE = c(alldrug_int.mae,
          alldrug_bach.ridge.mae),
  
  Adjusted_R2 = c(alldrug_int.adjr2,
                  alldrug_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = '"All Drugs" Model Comparison (Poverty*Education)')

```

### Heroin

```{r, eval=FALSE}

# data cleaning
data_heroin = od_data %>%
  select(heroin_rate, med_income, poverty, hs_grad) %>%
  drop_na()

x = model.matrix(heroin_rate ~ med_income + poverty*hs_grad,
                 data = data_heroin)[, -1]

y = data_heroin$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
heroin_int.mse = mean((actual - ridge.pred)^2)
heroin_int.rmse = sqrt(heroin_int.mse)
heroin_int.mae = mean(abs(actual - ridge.pred))

# R-squared
heroin_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

heroin_int.adjr2 = 1 - ((1 - heroin_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_int.mse, "\n")
cat("RMSE =", heroin_int.rmse, "\n")
cat("MAE =", heroin_int.mae, "\n")
cat("Adjusted R-squared =", heroin_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(heroin_int.rmse,
           heroin_hs.ridge.rmse),
  
  MAE = c(heroin_int.mae,
          heroin_hs.ridge.mae),
  
  Adjusted_R2 = c(heroin_int.adjr2,
                  heroin_hs.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Heroin Model Comparison (Poverty*Education)')

```

### Prescription Opioids

```{r, eval=FALSE}

# data cleaning
data_rxopioids = od_data %>%
  select(rxopioids_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(rxopioids_rate ~ med_income + poverty*bachelors,
                 data = data_rxopioids)[, -1]

y = data_rxopioids$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_int.mse = mean((actual - ridge.pred)^2)
rxopioids_int.rmse = sqrt(rxopioids_int.mse)
rxopioids_int.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_int.adjr2 = 1 - ((1 - rxopioids_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_int.mse, "\n")
cat("RMSE =", rxopioids_int.rmse, "\n")
cat("MAE =", rxopioids_int.mae, "\n")
cat("Adjusted R-squared =", rxopioids_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(rxopioids_int.rmse,
           rxopioids_bach.ridge.rmse),
  
  MAE = c(rxopioids_int.mae,
          rxopioids_bach.ridge.mae),
  
  Adjusted_R2 = c(rxopioids_int.adjr2,
                  rxopioids_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Prescription Opioids Model Comparison (Poverty*Education)')

```

### Cocaine

```{r, eval=FALSE}

# data cleaning
data_cocaine = od_data %>%
  select(cocaine_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(cocaine_rate ~ med_income + poverty*bachelors,
                 data = data_cocaine)[, -1]

y = data_cocaine$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_int.mse = mean((actual - ridge.pred)^2)
cocaine_int.rmse = sqrt(cocaine_int.mse)
cocaine_int.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_int.adjr2 = 1 - ((1 - cocaine_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", cocaine_int.mse, "\n")
cat("RMSE =", cocaine_int.rmse, "\n")
cat("MAE =", cocaine_int.mae, "\n")
cat("Adjusted R-squared =", cocaine_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(cocaine_int.rmse,
           cocaine_bach.ridge.rmse),
  
  MAE = c(cocaine_int.mae,
          cocaine_bach.ridge.mae),
  
  Adjusted_R2 = c(cocaine_int.adjr2,
                  cocaine_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Cocaine Model Comparison (Poverty*Education)')

```

### Methamphetamine

```{r, eval=FALSE}

# data cleaning
data_meth = od_data %>%
  select(meth_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(meth_rate ~ med_income + poverty*bachelors,
                 data = data_meth)[, -1]

y = data_meth$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
meth_int.mse = mean((actual - ridge.pred)^2)
meth_int.rmse = sqrt(meth_int.mse)
meth_int.mae = mean(abs(actual - ridge.pred))

# R-squared
meth_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

meth_int.adjr2 = 1 - ((1 - meth_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_int.mse, "\n")
cat("RMSE =", meth_int.rmse, "\n")
cat("MAE =", meth_int.mae, "\n")
cat("Adjusted R-squared =", meth_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(meth_int.rmse,
           meth_bach.ridge.rmse),
  
  MAE = c(meth_int.mae,
          meth_bach.ridge.mae),
  
  Adjusted_R2 = c(meth_int.adjr2,
                  meth_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Meth Model Comparison (Poverty*Education)')

```

### Benzodiazepines

```{r, eval=FALSE}

# data cleaning
data_benzo = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, hs_grad) %>%
  drop_na()

x = model.matrix(benzodiazepines_rate ~ med_income + poverty*hs_grad,
                 data = data_benzo)[, -1]

y = data_benzo$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_int.mse = mean((actual - ridge.pred)^2)
benzo_int.rmse = sqrt(benzo_int.mse)
benzo_int.mae = mean(abs(actual - ridge.pred))

# R2
benzo_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R2
n = length(actual)
p = ncol(x)

benzo_int.adjr2 = 1 - ((1 - benzo_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_int.mse, "\n")
cat("RMSE =", benzo_int.rmse, "\n")
cat("MAE =", benzo_int.mae, "\n")
cat("Adjusted R-squared =", benzo_int.adjr2, "\n")

```

```{r, eval=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(benzo_int.rmse,
           benzo_hs.ridge.rmse),
  
  MAE = c(benzo_int.mae,
          benzo_hs.ridge.mae),
  
  Adjusted_R2 = c(benzo_int.adjr2,
                  benzo_hs.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Benzodiazepines Model Comparison (Poverty*Education)')

```

# Research Question 3: Are there significant differences in overdose death rate between males and females?

## Results

```{r, eval=FALSE}

# data cleaning
gender_data = od_data %>%
  select(male_rate, female_rate) %>%
  drop_na()


# do males differ from females t-test
t_test_gender = t.test(gender_data$male_rate,
                       gender_data$female_rate,
                       paired = FALSE)

t_test_gender
colMeans(gender_data)

```

```{r, eval=FALSE}

gender_long = od_data %>%
  select(male_rate, female_rate) %>%
  pivot_longer(cols = everything(),
               names_to = "Sex",
               values_to = "Rate") %>%
  drop_na()

ggplot(gender_long, aes(x = Sex, y = Rate, fill = Sex)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(
    "female_rate" = "Female",
    "male_rate" = "Male"
  )) +
  labs(
    title = "Overdose Death Rates by Sex",
    x = "Sex",
    y = "Overdose Death Rate (per 100,000)"
  ) +
  theme(legend.position = "none")

```

# Research Question 4: Are there significant differences in overdose death rate between different age groups?

## Results

```{r, eval=FALSE}

age_long = od_data %>%
  select(age_15_24_rate,
         age_25_34_rate,
         age_35_44_rate,
         age_45_54_rate,
         age_55_64_rate,
         age_65plus_rate) %>%
  pivot_longer(cols = everything(),
               names_to = "age_group",
               values_to = "rate") %>%
  drop_na()

# Levene's test to see if variance is equal across groups
leveneTest(rate ~ age_group, data = age_long)

# assumption is violated, so do Welch's ANOVA
welch_age = oneway.test(rate ~ age_group,
                         data = age_long,
                         var.equal = FALSE)

welch_age

pairwise_age = pairwise.t.test(age_long$rate,
                                age_long$age_group,
                                p.adjust.method = "BH",
                                pool.sd = FALSE)

pairwise_age

```

```{r, eval=FALSE}

pairwise_table = as.data.frame(as.table(pairwise_age$p.value))

pairwise_table = na.omit(pairwise_table)

colnames(pairwise_table) = c("Group_1", "Group_2", "Adjusted_p_value")

pairwise_table$Group_1 = dplyr::recode(pairwise_table$Group_1,
  age_15_24_rate = "Ages 15-24",
  age_25_34_rate = "Ages 25-34",
  age_35_44_rate = "Ages 35-44",
  age_45_54_rate = "Ages 45-54",
  age_55_64_rate = "Ages 55-64",
  age_65plus_rate = "Ages 65+"
)

pairwise_table$Group_2 = dplyr::recode(pairwise_table$Group_2,
  age_15_24_rate = "Ages 15-24",
  age_25_34_rate = "Ages 25-34",
  age_35_44_rate = "Ages 35-44",
  age_45_54_rate = "Ages 45-54",
  age_55_64_rate = "Ages 55-64",
  age_65plus_rate = "Ages 65+"
)

# significance column
pairwise_table$Significant <- ifelse(
  pairwise_table$Adjusted_p_value < 0.05,
  "Yes",
  "No"
)

knitr::kable(
  pairwise_table,
  digits = 4,
  col.names = c("Age Group 1", "Age Group 2", "Adjusted p-value", "Significant?")
)
```

```{r, eval=FALSE}

ggplot(age_long, aes(x = age_group, y = rate, fill = age_group)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(
    "age_15_24_rate" = "Ages 15-24",
    "age_25_34_rate" = "Ages 25-34",
    "age_35_44_rate" = "Ages 35-44",
    "age_45_54_rate" = "Ages 45-54",
    "age_55_64_rate" = "Ages 55-64",
    "age_65plus_rate" = "Ages 65+"
  )) +
  labs(
    title = "Overdose Death Rates by Age Group",
    x = "Age Group",
    y = "Death Rate (per 100,000)"
  ) +
  theme(legend.position = "none")

```

# Residual Plots

## All Drugs

```{r, eval=FALSE}

# Residuals
res_alldrug = actual.alldrug - as.vector(ridge.pred.alldrug)

plot(ridge.pred.alldrug, res_alldrug,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residual Plot: All Drugs Model")
abline(h = 0,
       col = "red",
       lwd = 2)


qqnorm(res_alldrug,
       main = "QQ Plot: All Drugs Model")
qqline(res_alldrug,
       col = "red",
       lwd = 2)

```

## Heroin

```{r, eval=FALSE}

# Residuals
res_heroin = actual.heroin - as.vector(ridge.pred.heroin)

plot(ridge.pred.heroin, res_heroin,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residual Plot: Heroin Model")
abline(h = 0,
       col = "red",
       lwd = 2)


qqnorm(res_heroin,
       main = "QQ Plot: Heroin Model")
qqline(res_heroin,
       col = "red",
       lwd = 2)

```

## Prescription Opioids

```{r, eval=FALSE}

# Residuals
res_rxopioids = resid(rxopioids_model_bach) 

plot(predict(rxopioids_model_bach), res_rxopioids, 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residual Plot: Prescription Opioids Model") 
abline(h=0)


qqnorm(res_rxopioids,
       main = "QQ Plot: Prescription Opioids Model")
qqline(res_rxopioids, col = "red")

```

## Cocaine

```{r, eval=FALSE}

# Residuals
res_cocaine = resid(cocaine_model_bach) 

plot(predict(cocaine_model_bach), res_cocaine, 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residual Plot: Cocaine Model") 
abline(h=0)


qqnorm(res_cocaine,
       main = "QQ Plot: Cocaine Model")
qqline(res_cocaine, col = "red")

```

## Methamphetamine

```{r, eval=FALSE}

# Residuals
res_meth = actual.meth - as.vector(ridge.pred.meth)

plot(ridge.pred.meth, res_meth,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residual Plot: Methamphetamine Model")
abline(h = 0,
       col = "red",
       lwd = 2)


qqnorm(res_meth,
       main = "QQ Plot: Methamphetamine Model")
qqline(res_meth,
       col = "red",
       lwd = 2)

```

## Benzodiazepines

```{r, eval=FALSE}

# Residuals
res_benzo = resid(benzo_model_hs) 

plot(predict(benzo_model_hs), res_benzo, 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residual Plot: Benzodiazepines Model") 
abline(h=0)

qqnorm(res_benzo,
       main = "QQ Plot: Benzodiazepines Model")
qqline(res_benzo, col = "red")

```
