---
title: "Research Question 2"
---

```{r, message=FALSE, warning=FALSE, include=FALSE}

# load packages used

library(readxl)
library(tidyverse)
library(glmnet)
library(randomForest)
library(caret)
library(car)
library(knitr)
library(patchwork)
library(leaps)

# load data
od_data = read_excel("overdose_data.xlsx")

# change state and year to factor variables
od_data = od_data %>%
  mutate(
    state = factor(state),
    year = factor(year),
    med_income = med_income/1000
  )

# change all 9999 values to NA
od_data = od_data %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))


```

# Does the effect of poverty on overdose death rates depend on educational attainment of the population?

## Statistical Methods

For each drug category, we choose the ridge regression model containing either `hs_grad` or `bachelors` based on the results above and add an interaction term between `poverty` and the education predictor, then compare the RMSE, MAE, and adjusted $R^2$ of this model to those of the original model without the interaction term.

Ridge regression is used for this analysis because it is more robust to the collinearities among socioeconomic predictors. Using ridge regression allows the predictor variables to remain as consistent as possible across the different drug models since we can keep all socioeconomic variables in every model regardless of multicollinearities. Although linear regression models with variable selection achieved slightly better predictive performances for some drug categories, the ridge regression models performed similarly enough that the modest reduction in predictive accuracy was considered an acceptable tradeoff for improved consistency and comparability across models.

## Results

### All drugs

```{r, include=FALSE}

data_bach = od_data %>%
  select(alldrug_rate, med_income, poverty, bachelors) %>%
  drop_na()

# ridge bachelors

x = model.matrix(alldrug_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.alldrug = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.alldrug = y

# MSE, RMSE, MAE
alldrug_bach.ridge.mse = mean((actual.alldrug - ridge.pred.alldrug)^2)
alldrug_bach.ridge.rmse = sqrt(alldrug_bach.ridge.mse)
alldrug_bach.ridge.mae = mean(abs(actual.alldrug - ridge.pred.alldrug))

# R-squared
alldrug_bach.ridge.r2 = 1 - sum((actual.alldrug - ridge.pred.alldrug)^2) /
  sum((actual.alldrug - mean(actual.alldrug))^2)

# Adjusted R-squared
n = length(actual.alldrug)
p = ncol(x)

alldrug_bach.ridge.adjr2 = 1 - ((1 - alldrug_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_alldrug = od_data %>%
  select(alldrug_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(alldrug_rate ~ med_income + poverty*bachelors,
                 data = data_alldrug)[, -1]

y = data_alldrug$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
alldrug_int.mse = mean((actual - ridge.pred)^2)
alldrug_int.rmse = sqrt(alldrug_int.mse)
alldrug_int.mae = mean(abs(actual - ridge.pred))

# R-squared
alldrug_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

alldrug_int.adjr2 = 1 - ((1 - alldrug_int.r2) * (n - 1)) /
  (n - p - 1)

cat("MSE =", alldrug_int.mse, "\n")
cat("RMSE =", alldrug_int.rmse, "\n")
cat("MAE =", alldrug_int.mae, "\n")
cat("Adjusted R-squared =", alldrug_int.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(alldrug_int.rmse,
           alldrug_bach.ridge.rmse),
  
  MAE = c(alldrug_int.mae,
          alldrug_bach.ridge.mae),
  
  Adjusted_R2 = c(alldrug_int.adjr2,
                  alldrug_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = '"All Drugs" Model Comparison (Poverty*Education)')

```

### Heroin

```{r, include=FALSE}

data_hs = od_data %>%
  select(heroin_rate, med_income, poverty, hs_grad) %>%
  drop_na()

# ridge hs

x = model.matrix(heroin_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.heroin = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.heroin = y

# MSE, RMSE, MAE
heroin_hs.ridge.mse = mean((actual.heroin - ridge.pred.heroin)^2)
heroin_hs.ridge.rmse = sqrt(heroin_hs.ridge.mse)
heroin_hs.ridge.mae = mean(abs(actual.heroin - ridge.pred.heroin))

# R-squared
heroin_hs.ridge.r2 = 1 - sum((actual.heroin - ridge.pred.heroin)^2) /
  sum((actual.heroin - mean(actual.heroin))^2)

# Adjusted R-squared
n = length(actual.heroin)
p = ncol(x)

heroin_hs.ridge.adjr2 = 1 - ((1 - heroin_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_heroin = od_data %>%
  select(heroin_rate, med_income, poverty, hs_grad) %>%
  drop_na()

x = model.matrix(heroin_rate ~ med_income + poverty*hs_grad,
                 data = data_heroin)[, -1]

y = data_heroin$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
heroin_int.mse = mean((actual - ridge.pred)^2)
heroin_int.rmse = sqrt(heroin_int.mse)
heroin_int.mae = mean(abs(actual - ridge.pred))

# R-squared
heroin_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

heroin_int.adjr2 = 1 - ((1 - heroin_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_int.mse, "\n")
cat("RMSE =", heroin_int.rmse, "\n")
cat("MAE =", heroin_int.mae, "\n")
cat("Adjusted R-squared =", heroin_int.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(heroin_int.rmse,
           heroin_hs.ridge.rmse),
  
  MAE = c(heroin_int.mae,
          heroin_hs.ridge.mae),
  
  Adjusted_R2 = c(heroin_int.adjr2,
                  heroin_hs.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Heroin Model Comparison (Poverty*Education)')

```

### Prescription Opioids

```{r, include=FALSE}

data_bach = od_data %>%
  select(rxopioids_rate, med_income, poverty, bachelors) %>%
  drop_na()

# ridge bachelors

x = model.matrix(rxopioids_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_bach.ridge.mse = mean((actual - ridge.pred)^2)
rxopioids_bach.ridge.rmse = sqrt(rxopioids_bach.ridge.mse)
rxopioids_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_bach.ridge.adjr2 = 1 - ((1 - rxopioids_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_rxopioids = od_data %>%
  select(rxopioids_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(rxopioids_rate ~ med_income + poverty*bachelors,
                 data = data_rxopioids)[, -1]

y = data_rxopioids$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_int.mse = mean((actual - ridge.pred)^2)
rxopioids_int.rmse = sqrt(rxopioids_int.mse)
rxopioids_int.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_int.adjr2 = 1 - ((1 - rxopioids_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_int.mse, "\n")
cat("RMSE =", rxopioids_int.rmse, "\n")
cat("MAE =", rxopioids_int.mae, "\n")
cat("Adjusted R-squared =", rxopioids_int.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(rxopioids_int.rmse,
           rxopioids_bach.ridge.rmse),
  
  MAE = c(rxopioids_int.mae,
          rxopioids_bach.ridge.mae),
  
  Adjusted_R2 = c(rxopioids_int.adjr2,
                  rxopioids_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Prescription Opioids Model Comparison (Poverty*Education)')

```

### Cocaine

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_cocaine = od_data %>%
  select(cocaine_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(cocaine_rate ~ med_income + poverty*bachelors,
                 data = data_cocaine)[, -1]

y = data_cocaine$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_int.mse = mean((actual - ridge.pred)^2)
cocaine_int.rmse = sqrt(cocaine_int.mse)
cocaine_int.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_int.adjr2 = 1 - ((1 - cocaine_int.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, include=FALSE}

data_bach = od_data %>%
  select(cocaine_rate, med_income, poverty, bachelors) %>%
  drop_na()

# ridge bachelors

x = model.matrix(cocaine_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_bach.ridge.mse = mean((actual - ridge.pred)^2)
cocaine_bach.ridge.rmse = sqrt(cocaine_bach.ridge.mse)
cocaine_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_bach.ridge.adjr2 = 1 - ((1 - cocaine_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(cocaine_int.rmse,
           cocaine_bach.ridge.rmse),
  
  MAE = c(cocaine_int.mae,
          cocaine_bach.ridge.mae),
  
  Adjusted_R2 = c(cocaine_int.adjr2,
                  cocaine_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Cocaine Model Comparison (Poverty*Education)')

```

### Methamphetamine

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_meth = od_data %>%
  select(meth_rate, med_income, poverty, bachelors) %>%
  drop_na()

x = model.matrix(meth_rate ~ med_income + poverty*bachelors,
                 data = data_meth)[, -1]

y = data_meth$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
meth_int.mse = mean((actual - ridge.pred)^2)
meth_int.rmse = sqrt(meth_int.mse)
meth_int.mae = mean(abs(actual - ridge.pred))

# R-squared
meth_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

meth_int.adjr2 = 1 - ((1 - meth_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_int.mse, "\n")
cat("RMSE =", meth_int.rmse, "\n")
cat("MAE =", meth_int.mae, "\n")
cat("Adjusted R-squared =", meth_int.adjr2, "\n")

```

```{r, include=FALSE}

data_bach = od_data %>%
  select(meth_rate, med_income, poverty, bachelors) %>%
  drop_na()

# ridge bachelors

x = model.matrix(meth_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.meth = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.meth = y

# MSE, RMSE, MAE
meth_bach.ridge.mse = mean((actual.meth - ridge.pred.meth)^2)
meth_bach.ridge.rmse = sqrt(meth_bach.ridge.mse)
meth_bach.ridge.mae = mean(abs(actual.meth - ridge.pred.meth))

# R-squared
meth_bach.ridge.r2 = 1 - sum((actual.meth - ridge.pred.meth)^2) /
  sum((actual.meth - mean(actual.meth))^2)

# Adjusted R-squared
n = length(actual.meth)
p = ncol(x)

meth_bach.ridge.adjr2 = 1 - ((1 - meth_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*bachelors interaction",
            "No interaction term"
            ),
  
  RMSE = c(meth_int.rmse,
           meth_bach.ridge.rmse),
  
  MAE = c(meth_int.mae,
          meth_bach.ridge.mae),
  
  Adjusted_R2 = c(meth_int.adjr2,
                  meth_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Meth Model Comparison (Poverty*Education)')

```

### Benzodiazepines

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

# data cleaning
data_benzo = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, hs_grad) %>%
  drop_na()

x = model.matrix(benzodiazepines_rate ~ med_income + poverty*hs_grad,
                 data = data_benzo)[, -1]

y = data_benzo$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_int.mse = mean((actual - ridge.pred)^2)
benzo_int.rmse = sqrt(benzo_int.mse)
benzo_int.mae = mean(abs(actual - ridge.pred))

# R2
benzo_int.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R2
n = length(actual)
p = ncol(x)

benzo_int.adjr2 = 1 - ((1 - benzo_int.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_int.mse, "\n")
cat("RMSE =", benzo_int.rmse, "\n")
cat("MAE =", benzo_int.mae, "\n")
cat("Adjusted R-squared =", benzo_int.adjr2, "\n")

```

```{r, include=FALSE}

data_hs = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, hs_grad) %>%
  drop_na()

# ridge hs

x = model.matrix(benzodiazepines_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_hs.ridge.mse = mean((actual - ridge.pred)^2)
benzo_hs.ridge.rmse = sqrt(benzo_hs.ridge.mse)
benzo_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
benzo_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

benzo_hs.ridge.adjr2 = 1 - ((1 - benzo_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### compare

comparison_table = data.frame(
  Model = c("poverty*hs_grad interaction",
            "No interaction term"
            ),
  
  RMSE = c(benzo_int.rmse,
           benzo_hs.ridge.rmse),
  
  MAE = c(benzo_int.mae,
          benzo_hs.ridge.mae),
  
  Adjusted_R2 = c(benzo_int.adjr2,
                  benzo_hs.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Benzodiazepines Model Comparison (Poverty*Education)')

```

## Analysis of Results

For drugs overall, heroin, and methamphetamine, the models without an interaction term between poverty rate and educational attainment levels performed slightly better. Conversely, the models with interaction terms performed slightly better for cocaine, prescription opioids, and benzodiazepines. These improvements were modest, however, suggesting that the effect of poverty on overdose mortality rates is relatively consistent across different levels of educational attainment. In other words, while both poverty and education are independently related to overdose mortality, the relationship between poverty and overdose death rates does not appear to substantially strengthen or weaken depending on the education level of the population.

At the same time, the similar patterns observed among the cocaine, prescription opioids, and benzodiazepines models mirror the similarities identified in the analysis for Research Question 1. These recurring patterns support the likelihood that additional contributing factors beyond the socioeconomic variables included in this study play an important role in overdose mortality for these substances.
