---
title: "Research Question 3"
---

```{r, message=FALSE, warning=FALSE, include=FALSE}

# load packages used

library(readxl)
library(tidyverse)
library(glmnet)
library(randomForest)
library(caret)
library(car)
library(knitr)
library(patchwork)
library(leaps)

# load data
od_data = read_excel("overdose_data.xlsx")

# change state and year to factor variables
od_data = od_data %>%
  mutate(
    state = factor(state),
    year = factor(year),
    med_income = med_income/1000
  )

# change all 9999 values to NA
od_data = od_data %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))


```

# Are there significant differences in overdose death rate between males and females?

## Statistical Methods

We perform a Welch two sample t-test to compare overall overdose death rates between males and females. We set the significance level at 0.05.

## Results

The mean overdose death rate for males is approximately 46.26, and the mean overdose death rate for females is approximately 19.49. The t-test indicates a statistically significant difference in means ($p<2.2e-16$) with a 95% confidence interval of (23.03, 30.51).

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
gender_data = od_data %>%
  select(male_rate, female_rate) %>%
  drop_na()


# do males differ from females t-test
t_test_gender = t.test(gender_data$male_rate,
                       gender_data$female_rate,
                       paired = FALSE)

t_test_gender
colMeans(gender_data)

```

```{r, echo=FALSE}

gender_long = od_data %>%
  select(male_rate, female_rate) %>%
  pivot_longer(cols = everything(),
               names_to = "Sex",
               values_to = "Rate") %>%
  drop_na()

ggplot(gender_long, aes(x = Sex, y = Rate, fill = Sex)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(
    "female_rate" = "Female",
    "male_rate" = "Male"
  )) +
  labs(
    title = "Overdose Death Rates by Sex",
    x = "Sex",
    y = "Overdose Death Rate (per 100,000)"
  ) +
  theme(legend.position = "none")

```

## Analysis of Results

Males had significantly higher average overdose death rates than females, with the mean male overdose rate over twice the female rate. Possible explanations for this disparity are that males may be more likely to engage in high-risk substance-use behaviors, use larger quantities of drugs, or be less likely to seek treatment for substance-use disorders or mental health conditions. Zero is well outside the confidence interval for the difference in means, which further suggests that sex is an important demographic factor associated with overdose mortality, as it points to a practically meaningful, not just statistically significant, difference between sexes.
