---
title: "Research Question 1"
---

```{r, message=FALSE, warning=FALSE, include=FALSE}

# load packages used

library(readxl)
library(tidyverse)
library(glmnet)
library(randomForest)
library(caret)
library(car)
library(knitr)
library(patchwork)
library(leaps)


# load data
od_data = read_excel("overdose_data.xlsx")

# change state and year to factor variables
od_data = od_data %>%
  mutate(
    state = factor(state),
    year = factor(year),
    med_income = med_income/1000
  )

# change all 9999 values to NA
od_data = od_data %>%
  mutate(across(where(is.numeric), ~ na_if(., 9999)))

```

# How are median household income, poverty rate, and educational attainment related to drug overdose death rates in states, and do these relationships vary across different drugs?

## Statistical Methods

For each drug (including "all drugs"), we begin by creating two linear regression models containing `med_income`, `poverty`, and either `hs_grad`, or `bachelors` as predictors of the overdose death rate. We perform backward elimination on the models using 10-fold cross-validation

`med_income` was found to have a Variable Inflation Factor (VIF) greater than 5 in most of the models, indicating the (unsurprising) presence of (multi)collinearities between the socioeconomic variables. Thus, ridge regressions with 10-fold cross-validation are also performed using both initial sets of predictors, and the four models are compared by root mean squared error (RMSE), mean absolute error (MAE), and adjusted $R^2$. Baseline values for each of these statistics are included as well using a model that predicts the same overdose death rate for every observation. The baseline RMSE was computed using the mean as the predicted value, and the baseline MAE was computed using the median as the predicted value.

We choose the model with the highest adjusted $R^2$ and lowest RMSE and MAE for comparisons across drug categories. In cases of conflicting differences (i.e., lower adjusted $R^2$/higher RMSE but lower MAE), we choose the model with the lowest MAE due to the skewed nature of the overdose death rate distributions. Residual plots for the chosen model for each drug are included on the "Residual Plots" page.

We set the significance level at 0.05 for all tests.

## Results

### All drugs

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(alldrug_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$alldrug_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(alldrug_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
alldrug_model_hs = lm(alldrug_rate ~ med_income + poverty + hs_grad, 
                       data = data_hs)

summary(alldrug_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(alldrug_rate ~ med_income + poverty + hs_grad,
             data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$alldrug_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

alldrug_hs.mse = min(mean.cv.errors)
alldrug_hs.rmse = sqrt(alldrug_hs.mse)
alldrug_hs.mae = mean(cv.mae)
alldrug_hs.adjr2 = summary(alldrug_model_hs)$adj.r.squared

cat("Test MSE =", alldrug_hs.mse, "\n")
cat("Test RMSE =", alldrug_hs.rmse, "\n")
cat("Test MAE =", alldrug_hs.mae, "\n")
cat("Adjusted R-squared =", alldrug_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(alldrug_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
#                       data = data_bach, 
#                       method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$alldrug_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(alldrug_rate ~ med_income + poverty + bachelors, 
                         data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
alldrug_model_bach = lm(alldrug_rate ~ poverty + bachelors, 
                       data = data_bach)


summary(alldrug_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(alldrug_rate ~ poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$alldrug_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

alldrug_bach.mse = min(mean.cv.errors)
alldrug_bach.rmse = sqrt(alldrug_bach.mse)
alldrug_bach.mae = mean(cv.mae)
alldrug_bach.adjr2 = summary(alldrug_model_bach)$adj.r.squared

cat("Test MSE =", alldrug_bach.mse, "\n")
cat("Test RMSE =", alldrug_bach.rmse, "\n")
cat("Test MAE =", alldrug_bach.mae, "\n")
cat("Adjusted R-squared =", alldrug_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(alldrug_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
alldrug_hs.ridge.mse = mean((actual - ridge.pred)^2)
alldrug_hs.ridge.rmse = sqrt(alldrug_hs.ridge.mse)
alldrug_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
alldrug_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
                sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

alldrug_hs.ridge.adjr2 = 1 - ((1 - alldrug_hs.ridge.r2) * (n - 1)) /
                    (n - p - 1)


cat("MSE =", alldrug_hs.ridge.mse, "\n")
cat("RMSE =", alldrug_hs.ridge.rmse, "\n")
cat("MAE =", alldrug_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", alldrug_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(alldrug_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$alldrug_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.alldrug = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.alldrug = y

# MSE, RMSE, MAE
alldrug_bach.ridge.mse = mean((actual.alldrug - ridge.pred.alldrug)^2)
alldrug_bach.ridge.rmse = sqrt(alldrug_bach.ridge.mse)
alldrug_bach.ridge.mae = mean(abs(actual.alldrug - ridge.pred.alldrug))

# R-squared
alldrug_bach.ridge.r2 = 1 - sum((actual.alldrug - ridge.pred.alldrug)^2) /
  sum((actual.alldrug - mean(actual.alldrug))^2)

# Adjusted R-squared
n = length(actual.alldrug)
p = ncol(x)

alldrug_bach.ridge.adjr2 = 1 - ((1 - alldrug_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", alldrug_bach.ridge.mse, "\n")
cat("RMSE =", alldrug_bach.ridge.rmse, "\n")
cat("MAE =", alldrug_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", alldrug_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
### stats comparisons

# baseline model
alldrug_baseline.pred = mean(od_data$alldrug_rate, na.rm = TRUE)

alldrug_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


alldrug_baseline.rmse = sqrt(mean((od_data$alldrug_rate - alldrug_baseline.pred)^2,
                          na.rm = TRUE))

alldrug_baseline.mae = mean(abs(od_data$alldrug_rate - alldrug_baseline.pred.med),
                    na.rm = TRUE)

alldrug_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(alldrug_baseline.rmse,
           alldrug_hs.rmse,
           alldrug_bach.rmse,
           alldrug_hs.ridge.rmse,
           alldrug_bach.ridge.rmse),
  
  MAE = c(alldrug_baseline.mae,
          alldrug_hs.mae,
          alldrug_bach.mae,
          alldrug_hs.ridge.mae,
          alldrug_bach.ridge.mae),
  
  Adjusted_R2 = c(alldrug_baseline.adjr2,
                  alldrug_hs.adjr2,
                  alldrug_bach.adjr2,
                  alldrug_hs.ridge.adjr2,
                  alldrug_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = '"All Drugs" Model Comparison')

```

The chosen model is "ridge regression (bachelors)":

`alldrug_rate` = -38.70 - 0.06\*`med_income` + 3.09\*`poverty` + 1.10\*`bachelors`

### Heroin

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(heroin_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$heroin_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(heroin_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
heroin_model_hs = lm(heroin_rate ~ med_income + poverty + hs_grad, 
                      data = data_hs)

summary(heroin_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(heroin_rate ~ med_income + poverty + hs_grad,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$heroin_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

heroin_hs.mse = min(mean.cv.errors)
heroin_hs.rmse = sqrt(heroin_hs.mse)
heroin_hs.mae = mean(cv.mae)
heroin_hs.adjr2 = summary(heroin_model_hs)$adj.r.squared

cat("Test MSE =", heroin_hs.mse, "\n")
cat("Test RMSE =", heroin_hs.rmse, "\n")
cat("Test MAE =", heroin_hs.mae, "\n")
cat("Adjusted R-squared =", heroin_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(heroin_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$heroin_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(heroin_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
heroin_model_bach = lm(heroin_rate ~ med_income + poverty + bachelors, 
                        data = data_bach)


summary(heroin_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(heroin_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$heroin_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

heroin_bach.mse = min(mean.cv.errors)
heroin_bach.rmse = sqrt(heroin_bach.mse)
heroin_bach.mae = mean(cv.mae)
heroin_bach.adjr2 = summary(heroin_model_bach)$adj.r.squared

cat("Test MSE =", heroin_bach.mse, "\n")
cat("Test RMSE =", heroin_bach.rmse, "\n")
cat("Test MAE =", heroin_bach.mae, "\n")
cat("Adjusted R-squared =", heroin_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(heroin_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.heroin = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.heroin = y

# MSE, RMSE, MAE
heroin_hs.ridge.mse = mean((actual.heroin - ridge.pred.heroin)^2)
heroin_hs.ridge.rmse = sqrt(heroin_hs.ridge.mse)
heroin_hs.ridge.mae = mean(abs(actual.heroin - ridge.pred.heroin))

# R-squared
heroin_hs.ridge.r2 = 1 - sum((actual.heroin - ridge.pred.heroin)^2) /
  sum((actual.heroin - mean(actual.heroin))^2)

# Adjusted R-squared
n = length(actual.heroin)
p = ncol(x)

heroin_hs.ridge.adjr2 = 1 - ((1 - heroin_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_hs.ridge.mse, "\n")
cat("RMSE =", heroin_hs.ridge.rmse, "\n")
cat("MAE =", heroin_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", heroin_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(heroin_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$heroin_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
heroin_bach.ridge.mse = mean((actual - ridge.pred)^2)
heroin_bach.ridge.rmse = sqrt(heroin_bach.ridge.mse)
heroin_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
heroin_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

heroin_bach.ridge.adjr2 = 1 - ((1 - heroin_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", heroin_bach.ridge.mse, "\n")
cat("RMSE =", heroin_bach.ridge.rmse, "\n")
cat("MAE =", heroin_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", heroin_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### stats comparisons

# baseline model
heroin_baseline.pred = mean(od_data$heroin_rate, na.rm = TRUE)

heroin_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


heroin_baseline.rmse = sqrt(mean((od_data$heroin_rate - heroin_baseline.pred)^2,
                                  na.rm = TRUE))

heroin_baseline.mae = mean(abs(od_data$heroin_rate - heroin_baseline.pred.med),
                            na.rm = TRUE)

heroin_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(heroin_baseline.rmse,
           heroin_hs.rmse,
           heroin_bach.rmse,
           heroin_hs.ridge.rmse,
           heroin_bach.ridge.rmse),
  
  MAE = c(heroin_baseline.mae,
          heroin_hs.mae,
          heroin_bach.mae,
          heroin_hs.ridge.mae,
          heroin_bach.ridge.mae),
  
  Adjusted_R2 = c(heroin_baseline.adjr2,
                  heroin_hs.adjr2,
                  heroin_bach.adjr2,
                  heroin_hs.ridge.adjr2,
                  heroin_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Heroin Model Comparison')

```

The chosen model is "ridge regression (hs_grad)":

`heroin_rate` = -41.03 + 0.09\*`med_income` + 0.61\*`poverty` + 0.32\*`hs_grad`

### Prescription Opioids

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(rxopioids_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$rxopioids_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(rxopioids_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")

coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
rxopioids_model_hs = lm(rxopioids_rate ~ med_income + poverty, 
                     data = data_hs)

summary(rxopioids_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(rxopioids_rate ~ med_income + poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$rxopioids_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

rxopioids_hs.mse = min(mean.cv.errors)
rxopioids_hs.rmse = sqrt(rxopioids_hs.mse)
rxopioids_hs.mae = mean(cv.mae)
rxopioids_hs.adjr2 = summary(rxopioids_model_hs)$adj.r.squared

cat("Test MSE =", rxopioids_hs.mse, "\n")
cat("Test RMSE =", rxopioids_hs.rmse, "\n")
cat("Test MAE =", rxopioids_hs.mae, "\n")
cat("Adjusted R-squared =", rxopioids_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(rxopioids_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$rxopioids_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(rxopioids_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
rxopioids_model_bach = lm(rxopioids_rate ~ poverty + bachelors, 
                       data = data_bach)


summary(rxopioids_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(rxopioids_rate ~ poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$rxopioids_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

rxopioids_bach.mse = min(mean.cv.errors)
rxopioids_bach.rmse = sqrt(rxopioids_bach.mse)
rxopioids_bach.mae = mean(cv.mae)
rxopioids_bach.adjr2 = summary(rxopioids_model_bach)$adj.r.squared

cat("Test MSE =", rxopioids_bach.mse, "\n")
cat("Test RMSE =", rxopioids_bach.rmse, "\n")
cat("Test MAE =", rxopioids_bach.mae, "\n")
cat("Adjusted R-squared =", rxopioids_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(rxopioids_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_hs.ridge.mse = mean((actual - ridge.pred)^2)
rxopioids_hs.ridge.rmse = sqrt(rxopioids_hs.ridge.mse)
rxopioids_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_hs.ridge.adjr2 = 1 - ((1 - rxopioids_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_hs.ridge.mse, "\n")
cat("RMSE =", rxopioids_hs.ridge.rmse, "\n")
cat("MAE =", rxopioids_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", rxopioids_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(rxopioids_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$rxopioids_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
rxopioids_bach.ridge.mse = mean((actual - ridge.pred)^2)
rxopioids_bach.ridge.rmse = sqrt(rxopioids_bach.ridge.mse)
rxopioids_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
rxopioids_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

rxopioids_bach.ridge.adjr2 = 1 - ((1 - rxopioids_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", rxopioids_bach.ridge.mse, "\n")
cat("RMSE =", rxopioids_bach.ridge.rmse, "\n")
cat("MAE =", rxopioids_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", rxopioids_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### stats comparisons

# baseline model
rxopioids_baseline.pred = mean(od_data$rxopioids_rate, na.rm = TRUE)

rxopioids_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


rxopioids_baseline.rmse = sqrt(mean((od_data$rxopioids_rate - rxopioids_baseline.pred)^2,
                                 na.rm = TRUE))

rxopioids_baseline.mae = mean(abs(od_data$rxopioids_rate - rxopioids_baseline.pred.med),
                           na.rm = TRUE)

rxopioids_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(rxopioids_baseline.rmse,
           rxopioids_hs.rmse,
           rxopioids_bach.rmse,
           rxopioids_hs.ridge.rmse,
           rxopioids_bach.ridge.rmse),
  
  MAE = c(rxopioids_baseline.mae,
          rxopioids_hs.mae,
          rxopioids_bach.mae,
          rxopioids_hs.ridge.mae,
          rxopioids_bach.ridge.mae),
  
  Adjusted_R2 = c(rxopioids_baseline.adjr2,
                  rxopioids_hs.adjr2,
                  rxopioids_bach.adjr2,
                  rxopioids_hs.ridge.adjr2,
                  rxopioids_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Prescription Opioids Model Comparison')

```

The chosen model is "linear regression with variable selection (bachelors)":

`rxopioids_rate` = -3.51 + 0.40\*`poverty` + 0.09\*`bachelors`

### Cocaine

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(cocaine_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$cocaine_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(cocaine_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
cocaine_model_hs = lm(cocaine_rate ~ med_income + poverty + hs_grad, 
                     data = data_hs)

summary(cocaine_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(cocaine_rate ~ med_income + poverty + hs_grad,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$cocaine_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

cocaine_hs.mse = min(mean.cv.errors)
cocaine_hs.rmse = sqrt(cocaine_hs.mse)
cocaine_hs.mae = mean(cv.mae)
cocaine_hs.adjr2 = summary(cocaine_model_hs)$adj.r.squared

cat("Test MSE =", cocaine_hs.mse, "\n")
cat("Test RMSE =", cocaine_hs.rmse, "\n")
cat("Test MAE =", cocaine_hs.mae, "\n")
cat("Adjusted R-squared =", cocaine_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(cocaine_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$cocaine_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(cocaine_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
cocaine_model_bach = lm(cocaine_rate ~ med_income + bachelors, 
                       data = data_bach)


summary(cocaine_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(cocaine_rate ~ med_income + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$cocaine_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

cocaine_bach.mse = min(mean.cv.errors)
cocaine_bach.rmse = sqrt(cocaine_bach.mse)
cocaine_bach.mae = mean(cv.mae)
cocaine_bach.adjr2 = summary(cocaine_model_bach)$adj.r.squared

cat("Test MSE =", cocaine_bach.mse, "\n")
cat("Test RMSE =", cocaine_bach.rmse, "\n")
cat("Test MAE =", cocaine_bach.mae, "\n")
cat("Adjusted R-squared =", cocaine_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(cocaine_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_hs.ridge.mse = mean((actual - ridge.pred)^2)
cocaine_hs.ridge.rmse = sqrt(cocaine_hs.ridge.mse)
cocaine_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_hs.ridge.adjr2 = 1 - ((1 - cocaine_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", cocaine_hs.ridge.mse, "\n")
cat("RMSE =", cocaine_hs.ridge.rmse, "\n")
cat("MAE =", cocaine_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", cocaine_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(cocaine_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$cocaine_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
cocaine_bach.ridge.mse = mean((actual - ridge.pred)^2)
cocaine_bach.ridge.rmse = sqrt(cocaine_bach.ridge.mse)
cocaine_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
cocaine_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

cocaine_bach.ridge.adjr2 = 1 - ((1 - cocaine_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", cocaine_bach.ridge.mse, "\n")
cat("RMSE =", cocaine_bach.ridge.rmse, "\n")
cat("MAE =", cocaine_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", cocaine_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### stats comparisons

# baseline model
cocaine_baseline.pred = mean(od_data$cocaine_rate, na.rm = TRUE)

cocaine_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)

cocaine_baseline.rmse = sqrt(mean((od_data$cocaine_rate - cocaine_baseline.pred)^2,
                                 na.rm = TRUE))

cocaine_baseline.mae = mean(abs(od_data$cocaine_rate - cocaine_baseline.pred.med),
                           na.rm = TRUE)

cocaine_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(cocaine_baseline.rmse,
           cocaine_hs.rmse,
           cocaine_bach.rmse,
           cocaine_hs.ridge.rmse,
           cocaine_bach.ridge.rmse),
  
  MAE = c(cocaine_baseline.mae,
          cocaine_hs.mae,
          cocaine_bach.mae,
          cocaine_hs.ridge.mae,
          cocaine_bach.ridge.mae),
  
  Adjusted_R2 = c(cocaine_baseline.adjr2,
                  cocaine_hs.adjr2,
                  cocaine_bach.adjr2,
                  cocaine_hs.ridge.adjr2,
                  cocaine_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Cocaine Model Comparison')

```

The chosen model is "linear regression with variable selection (bachelors)":

`cocaine_rate` = -8.94 - 0.40\*`med_income` + 1.44\*`bachelors`

### Methamphetamine

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(meth_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$meth_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(meth_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
meth_model_hs = lm(meth_rate ~ poverty, 
                     data = data_hs)

summary(meth_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(meth_rate ~ poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$meth_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

meth_hs.mse = min(mean.cv.errors)
meth_hs.rmse = sqrt(meth_hs.mse)
meth_hs.mae = mean(cv.mae)
meth_hs.adjr2 = summary(meth_model_hs)$adj.r.squared

cat("Test MSE =", meth_hs.mse, "\n")
cat("Test RMSE =", meth_hs.rmse, "\n")
cat("Test MAE =", meth_hs.mae, "\n")
cat("Adjusted R-squared =", meth_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(meth_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$meth_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(meth_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
meth_model_bach = lm(meth_rate ~ med_income + poverty + bachelors, 
                       data = data_bach)


summary(meth_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(meth_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$meth_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

meth_bach.mse = min(mean.cv.errors)
meth_bach.rmse = sqrt(meth_bach.mse)
meth_bach.mae = mean(cv.mae)
meth_bach.adjr2 = summary(meth_model_bach)$adj.r.squared

cat("Test MSE =", meth_bach.mse, "\n")
cat("Test RMSE =", meth_bach.rmse, "\n")
cat("Test MAE =", meth_bach.mae, "\n")
cat("Adjusted R-squared =", meth_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(meth_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
meth_hs.ridge.mse = mean((actual - ridge.pred)^2)
meth_hs.ridge.rmse = sqrt(meth_hs.ridge.mse)
meth_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
meth_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

meth_hs.ridge.adjr2 = 1 - ((1 - meth_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_hs.ridge.mse, "\n")
cat("RMSE =", meth_hs.ridge.rmse, "\n")
cat("MAE =", meth_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", meth_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(meth_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$meth_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred.meth = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual.meth = y

# MSE, RMSE, MAE
meth_bach.ridge.mse = mean((actual.meth - ridge.pred.meth)^2)
meth_bach.ridge.rmse = sqrt(meth_bach.ridge.mse)
meth_bach.ridge.mae = mean(abs(actual.meth - ridge.pred.meth))

# R-squared
meth_bach.ridge.r2 = 1 - sum((actual.meth - ridge.pred.meth)^2) /
  sum((actual.meth - mean(actual.meth))^2)

# Adjusted R-squared
n = length(actual.meth)
p = ncol(x)

meth_bach.ridge.adjr2 = 1 - ((1 - meth_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", meth_bach.ridge.mse, "\n")
cat("RMSE =", meth_bach.ridge.rmse, "\n")
cat("MAE =", meth_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", meth_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### stats comparisons

# baseline model
meth_baseline.pred = mean(od_data$meth_rate, na.rm = TRUE)

meth_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


meth_baseline.rmse = sqrt(mean((od_data$meth_rate - meth_baseline.pred)^2,
                                 na.rm = TRUE))

meth_baseline.mae = mean(abs(od_data$meth_rate - meth_baseline.pred.med),
                           na.rm = TRUE)

meth_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(meth_baseline.rmse,
           meth_hs.rmse,
           meth_bach.rmse,
           meth_hs.ridge.rmse,
           meth_bach.ridge.rmse),
  
  MAE = c(meth_baseline.mae,
          meth_hs.mae,
          meth_bach.mae,
          meth_hs.ridge.mae,
          meth_bach.ridge.mae),
  
  Adjusted_R2 = c(meth_baseline.adjr2,
                  meth_hs.adjr2,
                  meth_bach.adjr2,
                  meth_hs.ridge.adjr2,
                  meth_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Methamphetamine Model Comparison')

```

The chosen model is "ridge regression (bachelors)":

`meth_rate` = 12.02 + 0.12\* `med_income` + 1.10\*`poverty` - 0.67\*`bachelors`

### Benzodiazepines

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning 
data_hs = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, hs_grad) %>%
  drop_na()

#backward elimination
#regfit_hs = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
#                       data = data_hs, method = "backward")


# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_hs), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
                        data = data_hs[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_hs[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_hs$benzodiazepines_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best_hs = regsubsets(benzodiazepines_rate ~ med_income + poverty + hs_grad, 
                         data = data_hs, nvmax = 3, method = "backward")
coef(reg.best_hs, which.min(mean.cv.errors))


# obtained model
benzo_model_hs = lm(benzodiazepines_rate ~ poverty, 
                     data = data_hs)

summary(benzo_model_hs)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(benzodiazepines_rate ~ poverty,
           data = data_hs[folds != j, ])
  
  pred = predict(fit, newdata = data_hs[folds == j, ])
  
  actual = data_hs$benzodiazepines_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

benzo_hs.mse = min(mean.cv.errors)
benzo_hs.rmse = sqrt(benzo_hs.mse)
benzo_hs.mae = mean(cv.mae)
benzo_hs.adjr2 = summary(benzo_model_hs)$adj.r.squared

cat("Test MSE =", benzo_hs.mse, "\n")
cat("Test RMSE =", benzo_hs.rmse, "\n")
cat("Test MAE =", benzo_hs.mae, "\n")
cat("Adjusted R-squared =", benzo_hs.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# data cleaning
data_bach = od_data %>%
  select(benzodiazepines_rate, med_income, poverty, bachelors) %>%
  drop_na()

#backward elimination
#regfit_bach = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
#                         data = data_bach, 
#                         method = "backward")

# backward elimination using 10-fold cross validation
k = 10
set.seed(1)

folds = sample(1:k, nrow(data_bach), replace = TRUE)

cv.errors = matrix(NA, k, 3, dimnames = list(NULL, paste(1:3)))

# function for prediction with regsubsets
predict.regsubsets = function(object, newdata, id, ...){
  form = as.formula(object$call[[2]])
  mat = model.matrix(form, newdata)
  coefi = coef(object, id = id)
  xvars = names(coefi)
  mat[, xvars] %*% coefi
}

for(j in 1:k){
  best.fit = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                        data = data_bach[folds != j, ],
                        nvmax = 3,
                        method = "backward")
  
  for(i in 1:3){
    pred = predict(best.fit, data_bach[folds == j, ], id = i)
    cv.errors[j, i] = mean((data_bach$benzodiazepines_rate[folds == j] - pred)^2)
  }
}

mean.cv.errors = apply(cv.errors, 2, mean)

plot(mean.cv.errors, type = "b")

# perform best subset selection and obtain the model
reg.best = regsubsets(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                      data = data_bach, nvmax = 3, method = "backward")
coef(reg.best, which.min(mean.cv.errors))


# obtained model
benzo_model_bach = lm(benzodiazepines_rate ~ med_income + poverty + bachelors, 
                       data = data_bach)


summary(benzo_model_bach)


# Test MSE, RMSE, MAE
cv.mae = rep(NA, k)

for(j in 1:k){
  fit = lm(benzodiazepines_rate ~ med_income + poverty + bachelors,
           data = data_bach[folds != j, ])
  
  pred = predict(fit, newdata = data_bach[folds == j, ])
  
  actual = data_bach$benzodiazepines_rate[folds == j]
  
  cv.mae[j] = mean(abs(actual - pred))
}

benzo_bach.mse = min(mean.cv.errors)
benzo_bach.rmse = sqrt(benzo_bach.mse)
benzo_bach.mae = mean(cv.mae)
benzo_bach.adjr2 = summary(benzo_model_bach)$adj.r.squared

cat("Test MSE =", benzo_bach.mse, "\n")
cat("Test RMSE =", benzo_bach.rmse, "\n")
cat("Test MAE =", benzo_bach.mae, "\n")
cat("Adjusted R-squared =", benzo_bach.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge hs

x = model.matrix(benzodiazepines_rate ~ med_income + poverty + hs_grad,
                 data = data_hs)[, -1]

y = data_hs$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_hs.ridge.mse = mean((actual - ridge.pred)^2)
benzo_hs.ridge.rmse = sqrt(benzo_hs.ridge.mse)
benzo_hs.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
benzo_hs.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

benzo_hs.ridge.adjr2 = 1 - ((1 - benzo_hs.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_hs.ridge.mse, "\n")
cat("RMSE =", benzo_hs.ridge.rmse, "\n")
cat("MAE =", benzo_hs.ridge.mae, "\n")
cat("Adjusted R-squared =", benzo_hs.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', include=FALSE}

# ridge bachelors

x = model.matrix(benzodiazepines_rate ~ med_income + poverty + bachelors,
                 data = data_bach)[, -1]

y = data_bach$benzodiazepines_rate

set.seed(1)

# 10-fold cross-validation ridge regression
cv.out.ridge = cv.glmnet(x, y, alpha = 0, nfolds = 10)

# best lambda
bestlam.ridge = cv.out.ridge$lambda.min

# final ridge model
ridge.mod = glmnet(x, y, alpha = 0, lambda = bestlam.ridge)

# coefficients
coef(ridge.mod)

# predictions using 10-fold CV-selected lambda
ridge.pred = predict(cv.out.ridge, s = "lambda.min", newx = x)

actual = y

# MSE, RMSE, MAE
benzo_bach.ridge.mse = mean((actual - ridge.pred)^2)
benzo_bach.ridge.rmse = sqrt(benzo_bach.ridge.mse)
benzo_bach.ridge.mae = mean(abs(actual - ridge.pred))

# R-squared
benzo_bach.ridge.r2 = 1 - sum((actual - ridge.pred)^2) /
  sum((actual - mean(actual))^2)

# Adjusted R-squared
n = length(actual)
p = ncol(x)

benzo_bach.ridge.adjr2 = 1 - ((1 - benzo_bach.ridge.r2) * (n - 1)) /
  (n - p - 1)


cat("MSE =", benzo_bach.ridge.mse, "\n")
cat("RMSE =", benzo_bach.ridge.rmse, "\n")
cat("MAE =", benzo_bach.ridge.mae, "\n")
cat("Adjusted R-squared =", benzo_bach.ridge.adjr2, "\n")

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}

### stats comparisons

# baseline model
benzo_baseline.pred = mean(od_data$benzodiazepines_rate, na.rm = TRUE)
benzo_baseline.pred.med = median(od_data$benzodiazepines_rate, na.rm = TRUE)


benzo_baseline.rmse = sqrt(mean((od_data$benzodiazepines_rate - benzo_baseline.pred)^2,
                                 na.rm = TRUE))

benzo_baseline.mae = mean(abs(od_data$benzodiazepines_rate - benzo_baseline.pred.med),
                           na.rm = TRUE)

benzo_baseline.adjr2 = 0


comparison_table = data.frame(
  Model = c("Baseline",
            "Linear regression with variable selection (hs_grad)",
            "Linear regression with variable selection (bachelors)",
            "Ridge regression (hs_grad)",
            "Ridge regression (bachelors)"),
  
  RMSE = c(benzo_baseline.rmse,
           benzo_hs.rmse,
           benzo_bach.rmse,
           benzo_hs.ridge.rmse,
           benzo_bach.ridge.rmse),
  
  MAE = c(benzo_baseline.mae,
          benzo_hs.mae,
          benzo_bach.mae,
          benzo_hs.ridge.mae,
          benzo_bach.ridge.mae),
  
  Adjusted_R2 = c(benzo_baseline.adjr2,
                  benzo_hs.adjr2,
                  benzo_bach.adjr2,
                  benzo_hs.ridge.adjr2,
                  benzo_bach.ridge.adjr2)
)

knitr::kable(comparison_table,
             digits = 3,
             caption = 'Benzodiazepines Model Comparison')

```

The chosen model is "linear regression with variable selection (hs_grad)":

`benzo_rate` = 0.84 + 0.24\*`poverty`

## Analysis of Results

Poverty rate was the most consistent predictor across all models, with higher poverty levels pointing to higher overdose death rates for all drugs except cocaine, with which it was not found to be significantly associated. The proportion of the population with a bachelor’s degree or higher was found to be a significant predictor of overdose death rates for all drugs except heroin and benzodiazepines, but the direction of the relationship was not consistent across models. Although higher proportions of the population with a bachelor’s degree were associated with lower rates of methamphetamine overdose deaths, they were associated with higher rates of death for prescription opioids, cocaine, benzodiazepines, and overall drugs. One possible explanation for this finding is that increased urbanization may contribute to both a greater prevalence and availability of certain drugs as well as to larger populations of college-educated professionals. In contrast, heroin overdose death rates were better explained by the proportion of the population with a high school degree or higher. Median household income also followed inconsistent patterns across the models, with higher median income associated with higher overdose death rates for heroin and methamphetamine but lower death rates for cocaine and all drugs combined.

Overall, the results indicate that socioeconomic conditions are significant predictors of overdose mortality rates, although the strength and nature of the relationships vary considerably between drug categories. The adjusted $R^2$ varied widely across models, from 0.60 in the cocaine model to 0.06 in the benzodiazepines model. The selected socioeconomic variables account for only about a third or less of the variance in the overdose death rate associated with every drug except cocaine, and they explain the least amount of variance in the overdose death rates of prescription opioids and benzodiazepines, the two drug groups in this study that are able to be obtained legally. This suggests that the factors, socioeconomic or otherwise, associated with overdose death rates may differ substantially across different drugs, and vulnerabilities to one substance may not transfer to another. Additionally, the similarities in performance between the ridge regression models and the models which underwent variable selection further highlight the complex relationship between socioeconomic factors and overdose mortality. The presence of multicollinearity among socioeconomic variables suggests that overdose death rates are influenced not by isolated socioeconomic characteristics, but by broader and highly interconnected social and economic conditions.
